Simple repression draft

### 2020-05-28 doc/section_04_bayesian_inference.tex

% !TEX root = ./bursty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference and model 
selection}
\label{section_04_bayesian_inference}
\subsection{Parameter inference for constitutive promoters}

@@

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior known
as the \textit{conjugate} prior for a Poisson likelihood. This type of priors
have nice properties to work with, in particular a conjugate prior allows us to
get a closed form for the posterior distribution $p(\lambda \mid D)$. In
particular it can be shown that for a Poisson distribution likelihood with its
Gamma distribution conjugate prior, the posterior distribution is also a Gamma
distribution ~\cite{Gelman2013}. In particular the two parameters $\alpha'$ and
$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
\frac{1}{N}\sum_k m_k$ for notational convenience, and $N$ is the number of
cells n our dataset. Furthermore given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to approximate the
Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
we have inferred our model's one parameter to a precision of order 1\%.

We remind the reader that we began this section claiming that the Poisson model
was ``wrong'' since it could not reproduce features of the data such as a Fano
factor $> 1$. The fact that we got such a narrow posterior distribution for our
parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
describe the data. What this means is that given the data $D$, the parameter 
$\lambda$ for our model is highly constrained to take a narrow range of values,
but a narrow posterior distribution does not speak to the goodness of fit of the
model. As we will see later in Figure~\ref{fig:constit_post_full} after
exploring the bursty promoter model, indeed there is not good correspondence
when contrasting the Poisson model with the experimental .

@@

Having found the model's posterior to be well-identified as with the Poisson
promoter, the next step is to compare both models with experimental data. To do
this for the case of the bursty promoter, for each of the parameter samples
shown in Figure~\ref{fig:constit_post_full}(A) we generated negative
bionomial-distributed mRNA counts. As MCMC samples parameter space
proportionally to the posterior distribution, this set of random samples span
the range of possible values that we would expect given the correspondence
between our theoretical model and the experimental data. A similar procedure can
be applied to the Poisson promoter. To compare so many samples with the actual
observed data, we can use empirical cumulative distribution functions (ECDF) of
the distribution quantiles. This representation is shown in
Figure~\ref{fig:constit_post_full}(B). In this example, the median for each
possible mRNA count for the Poisson distribution is shown as a dark green line,
while the lighter green contains 95\% of the randomly generated samples. This
way of representing the fit of the model to the data gives us a sense of the
range of data we might consider plausible, under the assumption that the model
is true. For this case, as we expected given our premise of the Poisson promoter
being wrong, it is quite obvious that the observed data, plotted in black. An 
equivalent plot for the bursty promoter model is shown in blue. Again the darker
tone shows the median, while the lighter color encompasses 95\% of the randomly
generated samples. Unlike the Poisson promoter model, the experimental ECDF
closely tracks the posterior predictive ECDF, indicating this model is actually
able to generate the observed data and increasing our confidence that this model
is at least not wrong.

>>>>

% !TEX root = ./bursty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference and model 
selection}
\label{section_04_bayesian_inference}

\textcolor{red}{In this section of the paper, we continue our program of providing
one complete description of the entire broad sweep of studies
that have been made in the context of the repressor-operator model,
dating all the way back to the original work of Jacob and Monod
and including the visionary quantitative work of M\"{u}ller-Hill and collaborators
and up to more recent studies.   In addition, the aim is to reconcile the 
equilibrium and non-equilibrium perspectives that have been brought
to bear on this problem.  To that end, we require a self-consistent means
of determining the parameters in these different theoretical analyses and
the current section does that, first by examining the constitutive promoter
and then turning to the full case of simple repression. }

\subsection{Parameter inference for constitutive promoters}

@@

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is known
as the \textit{conjugate} prior for a Poisson likelihood. This type of priors
have nice properties to work with, in particular a conjugate prior allows us to
get a closed form for the posterior distribution $p(\lambda \mid D)$. In
particular it can be shown that for a Poisson distribution likelihood with its
Gamma distribution conjugate prior, the posterior distribution is also a Gamma
distribution ~\cite{Gelman2013}. In particular the two parameters $\alpha'$ and
$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
\frac{1}{N}\sum_k m_k$ for notational convenience, and $N$ is the number of
cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to approximate the
Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
we have inferred our model's one parameter to a precision of order 1\%.

We remind the reader that we began this section claiming that the Poisson model
was ``wrong'' since it could not reproduce features of the data such as a Fano
factor $> 1$. The fact that we got such a narrow posterior distribution for our
parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
describe the data. What this means is that given the data $D$, the parameter 
$\lambda$ for our model is highly constrained to take a narrow range of values,
but a narrow posterior distribution does not speak to the goodness of fit of the
model. As we will see later in Figure~\ref{fig:constit_post_full} after
exploring the bursty promoter model, indeed there is not good correspondence
when contrasting the Poisson model with the experimental data.

@@

Having found the model's posterior to be well-identified as with the Poisson
promoter, the next step is to compare both models with experimental data. To do
this for the case of the bursty promoter, for each of the parameter samples
shown in Figure~\ref{fig:constit_post_full}(A) we generated negative
bionomial-distributed mRNA counts. As MCMC samples parameter space
proportionally to the posterior distribution, this set of random samples span
the range of possible values that we would expect given the correspondence
between our theoretical model and the experimental data. A similar procedure can
be applied to the Poisson promoter. To compare so many samples with the actual
observed data, we can use empirical cumulative distribution functions (ECDF) of
the distribution quantiles. This representation is shown in
Figure~\ref{fig:constit_post_full}(B). In this example, the median for each
possible mRNA count for the Poisson distribution is shown as a dark green line,
while the lighter green contains 95\% of the randomly generated samples. This
way of representing the fit of the model to the data gives us a sense of the
range of data we might consider plausible, under the assumption that the model
is true. For this case, as we expected given our premise of the Poisson promoter
being wrong, it is quite obvious that the observed data, plotted in black is not consistent with
the Poisson promoter model. An 
equivalent plot for the bursty promoter model is shown in blue. Again the darker
tone shows the median, while the lighter color encompasses 95\% of the randomly
generated samples. Unlike the Poisson promoter model, the experimental ECDF
closely tracks the posterior predictive ECDF, indicating this model is actually
able to generate the observed data and increasing our confidence that this model
is at least not wrong.
